---
title: "ESM244_Lab1"
author: "Allison Gracer"
date: "1/8/2022"
output: html_document
---

```{r setup, include = TRUE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(here)
library(sf)
library(tmap)
### install.packages("tmap")
### update.packages(ask = FALSE)
```

### Reading in the data

```{r}
sf_trees <- read_csv(here("data", "sf_trees", "sf_trees.csv"))
```

### Part 1: Wrangling and GGplot Review

**Example 1:** Find counts of observations by `legal_status` & wrangle a bit

```{r}
### method 1: group_by() %>%  summarize()
sf_trees %>% 
  group_by(legal_status) %>% 
  summarize(tree_count = n())

### method 2: different way plus a few new functions
top_5_status <- sf_trees %>% 
  count(legal_status) %>% 
  drop_na(legal_status) %>% 
  rename(tree_count = n) %>% 
  relocate(tree_count) %>% # easily re-order columns, this will relocate to front of my data
  slice_max(tree_count, n = 5) %>% 
  arrange(-tree_count) #- means arrange by descending, no - arrange by ascending or (desc(tree_count))
```

Make a graph of the top 5 from above

```{r}
ggplot(data = top_5_status, aes(x = fct_reorder(legal_status, tree_count), y = tree_count)) +
  geom_col(fill = "darkgreen") +
  labs(x = "Legal Status", y = "Tree Count") +
  coord_flip() +
  theme_minimal()
```

**Example 2:** Only keep observations where legal status is "Permitted Site" and caretaker is "MTA", and store as `permitted_data_df`

shift-control-c to comment/uncoment quickly (active or inactive code)
```{r}
# sf_trees$legal_status %>% unique()
# unique(sf_trees$caretaker)
permitted_data_df <- sf_trees %>% 
  filter(legal_status == c("Permitted Site") & caretaker == "MTA")
# filter(legal_status %in% c("Permitted Site", "Private")
```

**Example 3:** Only keep Blackwood Acacia trees, and then only keep columns `legal_status`, `date`, `latitude`, `longitude` and store as `blackwood_acacia_df`

```{r}
blackwood_acacia_df <- sf_trees %>% 
  filter(str_detect(species, "Blackwood Acacia")) %>% 
  select(legal_status, date, lat = latitude, lon = longitude)

### Make a little graph of locations
ggplot(data = blackwood_acacia_df, aes(x = lon, y = lat)) +
  geom_point(color = "darkgreen")
```

**Example 4:** Use `tidyr::seperate()` take on column and seperate it out

```{r}
sf_trees_sep <- sf_trees %>% 
  separate(species, into = c('spp_scientific', 'spp_common'), sep = ' :: ')
```

**Example 5:** use 'tidyr::unite()'

```{r}
ex_5 <- sf_trees %>% 
  unite('id_status', tree_id, legal_status, sep = '_COOL_')
```

### Part 2: make some maps

**Step 1:** convert lat/lon to spatial point, `st_as_sf()`

```{r}
blackwood_accacia_sf <- blackwood_acacia_df %>% 
  drop_na(lat, lon) %>% 
  st_as_sf(coords = c('lon', 'lat'))

### we need to tell R what the coordinate refernce system is
st_crs(blackwood_accacia_sf) <- 4326

```

